Model-based variable selection for regression with highly correlated variables.

### 2014-09-26 www/article.tex
up
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/correg@867 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
up
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/correg@869 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\paragraph{Qualitative illustration} Noting $\boldsymbol{\beta}^*$ the vector of coefficients of the same dimension as $\boldsymbol{\beta}$ with $\boldsymbol{\beta}_{J_r}^*=\boldsymbol{\beta}_{r}^*={\bf 0}$ and $\boldsymbol{\beta}_{J_f}^*=\boldsymbol{\beta}_{f}^*$, (\ref{eq:modexpl}) can be rewritten
\begin{equation}
	\boldsymbol{Y}{|\boldsymbol{X}_f,\boldsymbol{S}};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma}^2=\boldsymbol{X}_f\boldsymbol{\beta}_f^*+\boldsymbol{X}_r\boldsymbol{\beta}_r^*+\boldsymbol{\varepsilon}_Y^*.\label{eq:modexpl2}
\end{equation}
The {\sc ols} estimate of $\boldsymbol{\beta}^*$ is then given by
	\begin{equation}
		\hat{\boldsymbol{\beta}}^*_f=\left(\boldsymbol{X}_f'\boldsymbol{X}_f \right) ^{-1}\boldsymbol{X}_f'\boldsymbol{Y} \quad \mbox{and} \quad \hat{\boldsymbol{\beta}}^*_r = {\bf 0}.
	\end{equation}
As usual, the {\sc ols} estimate $\hat{\boldsymbol{\beta}}^*$ of $\boldsymbol{\beta}^*$ is unbiased but, however, contrary to $\hat{\boldsymbol{\beta}}$, it could be a biased estimate of $\boldsymbol{\beta}$ since
		\begin{equation}
			\mathbb{E}(\hat{\boldsymbol{\beta}}_f^*)=\boldsymbol{\beta}_f+\sum_{j =1}^{d_r}\beta_{J_r^j}\boldsymbol{\alpha}_j \quad \textrm{and}\quad \mathbb{E}(\hat{\boldsymbol{\beta}}_r^*)=\boldsymbol{0}.
		\end{equation}
In return, its variance could be reduced compared to this one of $\hat{\boldsymbol{\beta}}$ given in (\ref{eq:varOLS}) as soon as values of $\sigma_j$ are small enough (it means strong correlations in sub-regressions) as we can see in the following expression
		\begin{equation}
			\operatorname{Var}(\hat{\boldsymbol{\beta}}_f^*)= (\sigma^2_Y+\sum_{j =1}^{d_r}\sigma^2_{j}\beta_{J_r^j}^2)(\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1} \quad \textrm{and} \quad\operatorname{Var}(\hat{\boldsymbol{\beta}}_r^*)= \boldsymbol{0}. \label{eq:varOLS*}
		\end{equation}
Indeed, no correlations between covariates $\boldsymbol{X}_f$ imply that the matrix $\boldsymbol{X}_f' \boldsymbol{X}_f$ could be sufficiently better conditioned than the matrix $\boldsymbol{X}' \boldsymbol{X}$ involved in (\ref{eq:varOLS}) to balance the added variance $\sum_{j \in I_r}\sigma^2_{j}\beta_{j}^2$ in (\ref{eq:varOLS*}). This bias-variance trade off can be resumed by the Mean Squared Error (\textsc{mse}) associated to both estimates:
	\begin{eqnarray}
			\textsc{mse}(\hat{\boldsymbol{\beta}})&=& \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}),
			 \\
			\textsc{mse}(\hat{\boldsymbol{\beta}}^*)&=& \parallel\sum_{j =1}^{d_r}\sigma_{j}\beta_{J_r^j} \parallel_2^2 +\parallel \boldsymbol{\beta}_r\parallel^2_2 + (\sigma^2_Y+\sum_{j =1}^{d_r}\sigma^2_{j}\beta_{J_r^j}^2 ) \operatorname{Tr}((\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1}).
	\end{eqnarray}

@@

>>>>

\paragraph{Qualitative illustration} Noting $\boldsymbol{\beta}^*$ the vector of coefficients of the same dimension as $\boldsymbol{\beta}$ with $\boldsymbol{\beta}_{J_r}^*=\boldsymbol{\beta}_{r}^*={\bf 0}$ and $\boldsymbol{\beta}_{J_f}^*=\boldsymbol{\beta}_{f}^*$, (\ref{eq:modexpl}) can be rewritten
\begin{equation}
	\boldsymbol{Y}{|\boldsymbol{X}_f,\boldsymbol{S}};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma}^2=\boldsymbol{X}_f\boldsymbol{\beta}_f^*+\boldsymbol{X}_r\boldsymbol{\beta}_r^*+\boldsymbol{\varepsilon}_Y^*.\label{eq:modexpl2}
\end{equation}
The {\sc ols} estimate of $\boldsymbol{\beta}^*$ is then given by
	\begin{equation}
		\hat{\boldsymbol{\beta}}^*_f=\left(\boldsymbol{X}_f'\boldsymbol{X}_f \right) ^{-1}\boldsymbol{X}_f'\boldsymbol{Y} \quad \mbox{and} \quad \hat{\boldsymbol{\beta}}^*_r = {\bf 0}.
	\end{equation}
As usual, the {\sc ols} estimate $\hat{\boldsymbol{\beta}}^*$ of $\boldsymbol{\beta}^*$ is unbiased but, however, contrary to $\hat{\boldsymbol{\beta}}$, it could be a biased estimate of $\boldsymbol{\beta}$ since
		\begin{equation}
			\mathbb{E}(\hat{\boldsymbol{\beta}}_f^*)=\boldsymbol{\beta}_f+\sum_{j =1}^{d_r}\beta_{J_r^j}\boldsymbol{\alpha}_j^* \quad \textrm{and}\quad \mathbb{E}(\hat{\boldsymbol{\beta}}_r^*)=\boldsymbol{0}.
		\end{equation}
In return, its variance could be reduced compared to this one of $\hat{\boldsymbol{\beta}}$ given in (\ref{eq:varOLS}) as soon as values of $\sigma_j$ are small enough (it means strong correlations in sub-regressions) as we can see in the following expression
		\begin{equation}
			\operatorname{Var}(\hat{\boldsymbol{\beta}}_f^*)= (\sigma^2_Y+\sum_{j =1}^{d_r}\sigma^2_{j}\beta_{J_r^j}^2)(\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1} \quad \textrm{and} \quad\operatorname{Var}(\hat{\boldsymbol{\beta}}_r^*)= \boldsymbol{0}. \label{eq:varOLS*}
		\end{equation}
Indeed, no correlations between covariates $\boldsymbol{X}_f$ imply that the matrix $\boldsymbol{X}_f' \boldsymbol{X}_f$ could be sufficiently better conditioned than the matrix $\boldsymbol{X}' \boldsymbol{X}$ involved in (\ref{eq:varOLS}) to balance the added variance $\sum_{j \in I_r}\sigma^2_{j}\beta_{j}^2$ in (\ref{eq:varOLS*}). This bias-variance trade off can be resumed by the Mean Squared Error (\textsc{mse}) associated to both estimates:
	\begin{eqnarray}
			\textsc{mse}(\hat{\boldsymbol{\beta}})&=& \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}),
			 \\
			\textsc{mse}(\hat{\boldsymbol{\beta}}^*)&=& \parallel\sum_{j =1}^{d_r}\sigma_{j}\beta_{J_r^j} \parallel_2^2 +\parallel \boldsymbol{\beta}_r\parallel^2_2 + (\sigma^2_Y+\sum_{j =1}^{d_r}\sigma^2_{j}\beta_{J_r^j}^2 ) \operatorname{Tr}((\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1}).
	\end{eqnarray}

@@

		\section*{Appendix}\label{sectionident}
		\section{Identifiability}
		\subsection{Definition}
		We call identifiability:
\begin{eqnarray}
	\nexists (\boldsymbol{S},\tilde{\boldsymbol{S}}) \in \mathcal{S}_d\times \mathcal{S}_d \textrm{ with } \boldsymbol{S} \neq \tilde{\boldsymbol{S}} \textrm{ and }
	\mathbb{P}(\boldsymbol{X};\boldsymbol{S})=\mathbb{P}(\boldsymbol{X};\tilde{\boldsymbol{S}})
 \end{eqnarray}
 To avoid label-switching consideration, we suppose here (without loss of generality) that $\boldsymbol{J}_r$ is ordered by ascending order of the labels of the covariates.
	Hence identifiability is paired with the hypotheses we made on $ \mathcal{S}_d$. It is not sufficient to find a structure of linear sub-regression, the structure also has to verify hypotheses \ref{H1} to \ref{H3} (uncrossing rule + dependencies exhaustively described by the structure and then independence between the conditional response covariates). As a consequence, the covariance between two covariates is not null if and only if these covariates are linked by some sub-regressions.

		\subsection{Sufficient condition for identifiability}

\paragraph{Identifability criterion:} The model $\boldsymbol{S}$ is identifiable if
			\begin{equation}
				\forall j \in \{1,\dots,d_r\},  d_p^j>1\label{identcondition}.
			\end{equation}
That is to have at least two regressors in each sub-regression.\\

To prove the sufficiency of this condition for identifiability we rely on the following lemma.

\paragraph{Lemma:} With $\boldsymbol{X}$ and $\boldsymbol{S}$ following hypotheses \ref{H1} to \ref{H3}, covariance between two distinct covariates does differ from $0$ in only two cases:
\begin{enumerate}
	\item One of the two variables is a regressor of the other in a sub-regression	
	\begin{eqnarray}
		j \in \{1,\dots,d_r\}, i \in J_p^j \textrm{ then } \operatorname{cov}(\boldsymbol{X}^{i},\boldsymbol{X}^{J_r^j})\neq 0
	\end{eqnarray}
	\item Both variables are regressed by a common covariate in their respective sub-regression:
	\begin{eqnarray}
		\exists k \in J_f, \exists (i,j)\in \{1,\dots,d_r\}\times \{1,\dots,d_r\} \textrm{ with } i\neq j,&&  k\in J_p^i \textrm{ and } k\in J_p^j \\
		\textrm{ then } \operatorname{cov}(\boldsymbol{X}^{J_r^i},\boldsymbol{X}^{J_r^j})&\neq &0 \nonumber
	\end{eqnarray}
\end{enumerate}

\paragraph{proof of the lemma:}
The two cases lead immediately to non-zero covariance so we just look at other combinations of covariates.
\begin{itemize}
	\item if $\exists (i,j)\in \{1,\dots,d_r\}\times \{1,\dots,d_r\}, \operatorname{cov}(\boldsymbol{X}^{J_r^i},\boldsymbol{X}^{J_r^j})\neq 0$ then hypothesis \ref{H2} (uncrossing rule) guarantee that the two covariates are not in a same sub-regression so the covariance must come from the noises of the sub-regression but hypothesis \ref{H3} say that they are independent. The only remaining case is then the second case of the lemma: common covariate in the sub-regressions.
	\item if $(i,j)\in J_f\times J_f$ then  $\operatorname{cov}(\boldsymbol{X}^i,\boldsymbol{X}^j)=0 $ because covariates in $\boldsymbol{X}_f$ are orthogonal (by hypotheses \ref{H1} and \ref{H2}).
	\item if $j \in \{1,\dots,d_r\}, i \in J_f$ and $\operatorname{cov}(\boldsymbol{X}^{J_r^j},\boldsymbol{X}^i)\neq 0$ then $i\in J_p^j$ by hypotheses \ref{H1} and \ref{H3} and equation \ref{eq:H3}.
\end{itemize}
$\square$

\paragraph{proof of the identifiability criterion:} We suppose that \ref{identcondition} is verified and the model is not identifiable:
\begin{eqnarray}
	\exists (\boldsymbol{S},\tilde{\boldsymbol{S}}) \in \mathcal{S}_d\times \mathcal{S}_d \textrm{ with } \boldsymbol{S} \neq \tilde{\boldsymbol{S}} \textrm{ and }
	\mathbb{P}(\boldsymbol{X};\boldsymbol{S})=\mathbb{P}(\boldsymbol{X};\tilde{\boldsymbol{S}})
 \end{eqnarray}
 $\tilde{\boldsymbol{S}}=(\tilde{\boldsymbol{J}}_r,\tilde{\boldsymbol{J}}_p)$ contains $\tilde{d}_r$ sub-regressions and is characterized by $\tilde{\boldsymbol{J}}_r=(\tilde{J}_r^1,\dots,\tilde{J}_r^{\tilde{d}_r}),\tilde{\boldsymbol{J}}_p=(\tilde{J}_p^1,\dots,\tilde{J}_p^{\tilde{d}_r})$.\\
Because $\boldsymbol{S}\neq \tilde{\boldsymbol{S}} $ we have $\boldsymbol{J}_r\neq \tilde{\boldsymbol{J}}_r$ or $\boldsymbol{J}_p\neq \tilde{\boldsymbol{J}}_p$.
\begin{itemize}
\item If $\boldsymbol{J}_r = \tilde{\boldsymbol{J}}_r$ and $\boldsymbol{S}\neq \tilde{\boldsymbol{S}}$ then one sub-regression as a predictor that stands only in one of the two structures. We suppose (without loss of generality) that $\exists j \in \{1,\dots,d_r\}$ for which
$\exists i \in J_p^j$ with $i \notin \tilde{J}_p^j$ so $\operatorname{cov}_{S}(\boldsymbol{X}^{J_r^j},\boldsymbol{X}^{i})\neq 0$ and $\operatorname{cov}_{\tilde{S}}(\boldsymbol{X}^{J_r^j},\boldsymbol{X}^{i})= 0$ (from the lemma) so the two structure don't give the same joint distribution, leading to a contradiction.
\item If $\boldsymbol{J}_r\neq \tilde{\boldsymbol{J}}_r$ then one of the two models has a sub-regression that is not in the other. We suppose (without loss of generality) that
 $\exists J_r^j \in J_r$ with $J_r^j \notin \tilde{J}_r$ then $J_r^j\in \tilde{J_f}$ (recall $J_f=\{1,\dots,d \}\setminus J_r$). We note that $J_r^j \in J_r$ means $\exists k_1\neq k_2, \{k_1,k_2 \}\subset J_p^j \subset J_f$. Then $\operatorname{cov}_{\boldsymbol{S}}(\boldsymbol{X}^{J_r^j},\boldsymbol{X}^{k_1})\neq 0$ and $\operatorname{cov}_{\boldsymbol{S}}(\boldsymbol{X}^{J_r^j},\boldsymbol{X}^{k_2})\neq 0 $ so by the lemma $k_1$ and $k_2$ are responses variables in $\tilde{\boldsymbol{S}}$: $\exists (l_1,l_2) \in \{1,\dots,\tilde{d}_r \}\times \{1,\dots,\tilde{d}_r \}, \tilde{J}_r^{l_1}=k_1,\tilde{J}_r^{l_2}=k_2$ and $J_r^j$ is a regressor of $k_1$ and $k_2$: $J_r^j \in J_p^{l_1}, J_r^j \in J_p^{l_2}$ thus $\operatorname{cov}_{\tilde{\boldsymbol{S}}}(\boldsymbol{X}^{k_1},\boldsymbol{X}^{k_2})\neq 0$ that is not possible because $ \{k_1,k_2 \}\subset J_p^j \subset J_f$ and covariates in $\boldsymbol{X}^{J_f}$ are orthogonal by hypotheses.
\end{itemize}

Finally, condition \ref{identcondition} is sufficient for identifiability of $\boldsymbol{S}$.
  	$\square$

\paragraph{Remark:} Because sub-regressions with at least two regressors are identifiable, 
the only non-identifiable sub-regressions could be those with only one regressor, leading only to pairwise correlations that can be seen directly in the correlation matrix. Such sub-regression can be permuted without any impact on interpretation so such trivial sub-regression are not a problem even if they may occur with real datasets. One more thing: exact sub-regression with at least two sub-regressors are identifiable with our hypotheses.
